Fast Random Integer Generation in an Interval

### 2024-02-13 Presentation/unbiased.tex
some fixes

\subsection{The OpenBSD Algorithm}\label{sec:2.1}
\begin{frame}{The OpenBSD Algorithm}
    \pause 
    \begin{itemize}[<+->]
        \item We shift the \ER{rejection interval} to the left:
        \onslide<+->{\begin{align*}\everymath{} % Allows specific coloring
            \overbrace{\ER{\underbrace{0,1,\ldots,\RNW - 1}_{\text{$\RNW$ values}}},\underbrace{\overbrace{\RNW,\ldots,n - 1,0,\ldots,\RNW - 1}^{\text{$n$ values}},\ldots,\overbrace{\RNW,\ldots,n - 1,0,\ldots,\RNW - 1}^{\text{$n$ values}}}_{\text{$\left(2^W \div n\right) \cdot n$ values}}}^{\text{$2^W$ values}}
        \end{align*}}
        \item Algorithm: \begin{itemize}[<+->]
            \item Generate a uniform random number $x \in [0,2^W)$ until $x \geq \RNW$
            \item Return $x \Mod n$
        \end{itemize} 
    \end{itemize}

@@

\begin{frame}{The OpenBSD Algorithm}
    Algorithm: \begin{itemize}
        \item Generate a uniform random number $x \in [0,2^W)$ until $x \geq \RNW$
        \item Return $x \Mod n$
    \end{itemize} 
    \pause 
    \smallskip
    \begin{block}{Efficiency}<+->
        \onslide<+->{We require $2$ integer division operations:} \begin{itemize}[<+->]
            \item one for computing $\RNW$
            \item and one for computing $x \Mod n$.
        \end{itemize}
    \end{block}
    \vspace*{0.5cm} 
    \onslide<+->{Can we do better?}
\end{frame}

@@

\subsection{The Bitmask Algorithm}\label{sec:2.3}
\begin{frame}{The Bitmask Algorithm}
    \pause 
    \begin{itemize}[<+->]
        \item Consider the \GB{binary} representation of $n$: \onslide<+->{\begin{align*}\everymath{}
            n \quad \stackrel{\text{binary}}{\longrightarrow} \qquad \underbrace{
                \overbrace{0}^{
                    \hidewidth 2^{W - 1} \hidewidth 
                },\ldots,0
            }_{\text{only $0$'s}},\ER{\overbrace{1}^{
                 2^{\lfloor\log_2n\rfloor}
            }},\underbrace{
                \nicefrac{0}{1},\ldots,\overbrace{
                    \nicefrac{0}{1}
                }^{ 2^1 },\overbrace{
                    \nicefrac{0}{1}
                }^{ 2^0 }
            }_{\text{series of $0$'s and $1$'s}}
        \end{align*}} 
        \item Every number $x \leq n$ only needs the last $\lfloor\log_2n\rfloor + 1$ bits
        \item Get these bits with a bitwise-\textsc{And} with \begin{align*}\everymath{}
            2^{\lfloor\log_2n\rfloor + 1} - 1 \quad \stackrel{\text{binary}}{\longrightarrow} \qquad \underbrace{
                \overbrace{0}^{
                    \hidewidth 2^{W - 1} \hidewidth 
                },\ldots,0
            }_{\text{only $0$'s}},\underbrace{\ER{\overbrace{1}^{
                 2^{\lfloor\log_2n\rfloor}
            }},1,\ldots,\overbrace{
                    1
                }^{ 2^1 },\overbrace{
                    1
                }^{ 2^0 }
            }_{\text{only $1$'s}}
        \end{align*}
    \end{itemize}
\end{frame}

\begin{frame}{The Bitmask Algorithm}
    \pause 
    \begin{itemize}[<+->]
        \item How can we compute $2^{\lfloor\log_2n\rfloor + 1}$? 
        \item \GB{Count} the number $\ell$ of \GB{leading $0$'s}! \onslide<+->{\begin{align*}\everymath{}
            n \quad \stackrel{\text{binary}}{\longrightarrow} \qquad \underbrace{
                \overbrace{0}^{
                    \hidewidth 2^{W - 1} \hidewidth 
                },\ldots,0
            }_{\text{$\ell$ $0$'s}},\ER{\overbrace{1}^{
                 2^{\lfloor\log_2n\rfloor}
            }},\underbrace{
                \nicefrac{0}{1},\ldots,\overbrace{
                    \nicefrac{0}{1}
                }^{ 2^1 },\overbrace{
                    \nicefrac{0}{1}
                }^{ 2^0 }
            }_{\text{series of $0$'s and $1$'s}}
        \end{align*}} 
        \item $\ER{\lfloor\log_2n\rfloor} = W - \ell - 1$ \onslide<+->{$\qquad \longrightarrow \qquad 2^{\lfloor\log_2n\rfloor + 1} = 1 \ll \left(W - \ell\right)$}
        \item Algorithm: \begin{itemize}[<+->]
            \item[(1)] Compute $\ell$ and $\mathcal{M} = 2^{\lfloor\log_2n\rfloor + 1} - 1$
            \item[(2)] Draw $x \in [0,2^W)$ and compute $b = x \And \mathcal{M}$
            \item[(3)] Return $b$ if $b < n$ else goto \textbf{\GB{(2)}}
        \end{itemize}
    \end{itemize}
\end{frame}

\begin{frame}{The Bitmask Algorithm}
    Algorithm: \begin{itemize}
        \item[(1)] Compute $\ell$ and $\mathcal{M} = 2^{\lfloor\log_2n\rfloor + 1} - 1$
        \item[(2)] Draw $x \in [0,2^W)$ and compute $b = x \And \mathcal{M}$
        \item[(3)] Return $b$ if $b < n$ else goto \textbf{\GB{(2)}}
    \end{itemize}
    \pause 
    \smallskip
    \begin{block}{Efficiency}<+->
        \begin{itemize}[<+->]
            \item $b$ at most $\mathcal{M} = 2^{\lfloor\log_2n\rfloor + 1} - 1 < 2n$ \onslide<+->{$\quad \longrightarrow \quad$ success probability at least $\approx \frac{1}{2}$}
            \item At most $2$ rounds in expectation
            \item \GB{No} integer division at all
            \item Computation of \GB{leading $0$'s} requires \GB{\texttt{clz}} instruction/algorithm 
            \item Roughly as expensive as a \GB{\texttt{div}} instruction 
        \end{itemize}
    \end{block}
\end{frame}

>>>>

\subsection{The OpenBSD Algorithm}\label{sec:2.1}
\begin{frame}{The OpenBSD Algorithm}
    \pause 
    \begin{itemize}[<+->]
        \item Shift the \ER{rejection interval} to the left:
        \onslide<+->{\begin{align*}\everymath{} % Allows specific coloring
            \overbrace{\ER{\underbrace{0,1,\ldots,\RNW - 1}_{\text{$\RNW$ values}}},\underbrace{\overbrace{\RNW,\ldots,n - 1,0,\ldots,\RNW - 1}^{\text{$n$ values}},\ldots,\overbrace{\RNW,\ldots,n - 1,0,\ldots,\RNW - 1}^{\text{$n$ values}}}_{\text{$\left(2^W \div n\right) \cdot n$ values}}}^{\text{$2^W$ values}}
        \end{align*}}
        \item Algorithm: \begin{itemize}[<+->]
            \item Generate a uniform random number $x \in [0,2^W)$ until $x \geq \RNW$
            \item Return $x \Mod n$
        \end{itemize} 
    \end{itemize}

@@

\begin{frame}{The OpenBSD Algorithm}
    Algorithm: \begin{itemize}
        \item Generate a uniform random number $x \in [0,2^W)$ until $x \geq \RNW$
        \item Return $x \Mod n$
    \end{itemize} 
    \pause 
    \vspace*{0.5cm}
    \begin{block}{Efficiency}<+->
        \onslide<+->{We require $2$ integer division operations:} \begin{itemize}[<+->]
            \item one for computing $\RNW$
            \item and one for computing $x \Mod n$.
        \end{itemize}
    \end{block}
\end{frame}

@@

\subsection{The Bitmask Algorithm}\label{sec:2.3}
\begin{frame}{The Bitmask Algorithm}
    \pause 
    \begin{itemize}[<+->]
        \item Consider the \GB{binary} representation of $n$: \onslide<+->{\begin{align*}\everymath{}
            n \quad \xrightarrow{\text{binary}} \qquad \underbrace{
                \overbrace{0}^{
                    \hidewidth 2^{W - 1} \hidewidth 
                },\ldots,0
            }_{\text{only $0$'s}},\ER{\overbrace{1}^{
                 2^{\lfloor\log_2n\rfloor}
            }},\underbrace{
                \nicefrac{0}{1},\ldots,\overbrace{
                    \nicefrac{0}{1}
                }^{ 2^1 },\overbrace{
                    \nicefrac{0}{1}
                }^{ 2^0 }
            }_{\text{series of $0$'s and $1$'s}}
        \end{align*}} 
        \item Every number $x \leq n$ only needs the last $\lfloor\log_2n\rfloor + 1$ bits
        \item Get these bits with a bitwise-\textsc{And} with \begin{align*}\everymath{}
            2^{\lfloor\log_2n\rfloor + 1} - 1 \quad \xrightarrow{\text{binary}} \qquad \underbrace{
                \overbrace{0}^{
                    \hidewidth 2^{W - 1} \hidewidth 
                },\ldots,0
            }_{\text{only $0$'s}},\underbrace{\ER{\overbrace{1}^{
                 2^{\lfloor\log_2n\rfloor}
            }},1,\ldots,\overbrace{
                    1
                }^{ 2^1 },\overbrace{
                    1
                }^{ 2^0 }
            }_{\text{only $1$'s}}
        \end{align*}
    \end{itemize}
\end{frame}

\begin{frame}{The Bitmask Algorithm}
    \pause 
    \begin{itemize}[<+->]
        \item How can we compute $2^{\lfloor\log_2n\rfloor + 1}$? 
        \item \GB{Count} the number $\ell$ of \GB{leading $0$'s}! \onslide<+->{\begin{align*}\everymath{}
            n \quad \xrightarrow{\text{binary}} \qquad \underbrace{
                \overbrace{0}^{
                    \hidewidth 2^{W - 1} \hidewidth 
                },\ldots,0
            }_{\text{$\ell$ $0$'s}},\ER{\overbrace{1}^{
                 2^{\lfloor\log_2n\rfloor}
            }},\underbrace{
                \nicefrac{0}{1},\ldots,\overbrace{
                    \nicefrac{0}{1}
                }^{ 2^1 },\overbrace{
                    \nicefrac{0}{1}
                }^{ 2^0 }
            }_{\text{series of $0$'s and $1$'s}}
        \end{align*}} 
        \item $\ER{\lfloor\log_2n\rfloor} = W - \ell - 1$ \onslide<+->{$\qquad \longrightarrow \qquad 2^{\lfloor\log_2n\rfloor + 1} = 1 \ll \left(W - \ell\right)$}
        \item Algorithm: \begin{itemize}[<+->]
            \item[(1)] Compute $\ell$ and $\mathcal{M} = 2^{\lfloor\log_2n\rfloor + 1} - 1$
            \item[(2)] Draw $x \in [0,2^W)$ and compute $b = x \And \mathcal{M}$
            \item[(3)] Return $b$ if $b < n$ else goto \textbf{\GB{(2)}}
        \end{itemize}
    \end{itemize}
\end{frame}

\begin{frame}{The Bitmask Algorithm}
    Algorithm: \begin{itemize}
        \item[(1)] Compute $\ell$ and $\mathcal{M} = 2^{\lfloor\log_2n\rfloor + 1} - 1$
        \item[(2)] Draw $x \in [0,2^W)$ and compute $b = x \And \mathcal{M}$
        \item[(3)] Return $b$ if $b < n$ else goto \textbf{\GB{(2)}}
    \end{itemize}
    \pause 
    \smallskip
    \begin{block}{Efficiency}<+->
        \begin{itemize}[<+->]
            \item $b$ at most $\mathcal{M} = 2^{\lfloor\log_2n\rfloor + 1} - 1 < 2n$ \onslide<+->{$\quad \longrightarrow \quad$ success probability at least $\approx \frac{1}{2}$}
            \item At most $\approx 2$ rounds in expectation
            \item \GB{No} integer division at all
            \item Computation of \GB{leading $0$'s} requires \GB{\texttt{clz}} instruction/algorithm 
            \item Roughly as expensive as a \GB{\texttt{div}} instruction 
        \end{itemize}
    \end{block}
\end{frame}
